---
title: "How to use qgcomp package"
author: "Alexander Keil"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use qgcomp package}
  %\VignetteEngine{knitr::knitr}
  \usepackage[utf8]{inputenc}
---

## Introduction

## How to use the `qgcomp` package

### Example 1

Here are some observations from  `wqs_data` dataset from the from the package `gQWS`. 
These data will be used as an ongoing example to demonstrate features of the
`qgcomp` package. Parallel examples can be seen in the `gQWS` help files.

```{r, echo=FALSE, results='markup', message=FALSE}
library(gWQS)
library(qgcomp)
library(knitr)
head(wqs_data[, c(37, 36, 35, 1:34)], 10)
```

qgcomp with a continuous outcome:
This script calls a qgcomp model for a continuous outcome using the function `qgcomp.noboot`.
Compare with call to `gqws` from the package `gQWS`.

```{r, results='markup', fig.show='hold', fig.height=5, fig.width=5, cache=TRUE}

# we save the names of the mixture variables in the variable "Xnm"
data("wqs_data", package="gWQS")

Xnm <- c("log_LBX074LA", "log_LBX099LA", "log_LBX105LA", "log_LBX118LA",
"log_LBX138LA", "log_LBX153LA", "log_LBX156LA", "log_LBX157LA", "log_LBX167LA",
"log_LBX170LA", "log_LBX180LA", "log_LBX187LA", "log_LBX189LA", "log_LBX194LA",
"log_LBX196LA", "log_LBX199LA", "log_LBXD01LA", "log_LBXD02LA", "log_LBXD03LA",
"log_LBXD04LA", "log_LBXD05LA", "log_LBXD07LA", "log_LBXF01LA", "log_LBXF02LA",
"log_LBXF03LA", "log_LBXF04LA", "log_LBXF05LA", "log_LBXF06LA", "log_LBXF07LA",
"log_LBXF08LA", "log_LBXF09LA", "log_LBXPCBLA", "log_LBXTCDLA", "log_LBXHXCLA")


# Run the model and save the results in the variable "results"
results <- qgcomp.noboot(y~.,dat=wqs_data[,c(Xnm, 'y')], family=gaussian())


# Compare a qgcomp.noboot fit:
gcompmod <- qgcomp.noboot(disease_state~., expnms=Xnm, data = wqs_data[,c(Xnm, 'disease_state')], family=binomial(), q=4)

# and a qgcomp.boot fit:
gcompmod2 <- qgcomp.boot(disease_state~., expnms=Xnm, data = wqs_data[,c(Xnm, 'disease_state')], family=binomial(), q=4, B=200)


# with a gwqs fit:
suppressWarnings(wqsmod <- gwqs(disease_state ~ 1, mix_name = Xnm, data = wqs_data, q = 4, validation = 0.6, b = 3, b1_pos = F, b1_constr = F, family='binomial', seed=125))
     

# WQS fit (with reduced number of bootstraps to save time)
wqsmod$final_weights
wqsmod$fit
# conditional OR fit, similar to WQS
gcompmod
# population average RR with bootstrap confidence intervals
gcompmod2


```

### Example 2

In the following code we run an adjusted linear regression (`family = "gaussian"`)...

```{r, results='markup', fig.show='hold', fig.height=5, fig.width=5, cache=TRUE}

gcompmod <- qgcomp.noboot(y~sex+log_LBX074LA+log_LBX099LA+log_LBX105LA+log_LBX118LA+log_LBX138LA+log_LBX153LA+
                           log_LBX156LA+log_LBX157LA+log_LBX167LA+log_LBX170LA+log_LBX180LA+log_LBX187LA+log_LBX189LA+
                           log_LBX194LA+log_LBX196LA+log_LBX199LA+log_LBXD01LA+log_LBXD02LA+log_LBXD03LA+log_LBXD04LA+
                           log_LBXD05LA+log_LBXD07LA+log_LBXF01LA+log_LBXF02LA+log_LBXF03LA+log_LBXF04LA+log_LBXF05LA+
                           log_LBXF06LA+log_LBXF07LA+log_LBXF08LA+log_LBXF09LA+log_LBXPCBLA+log_LBXTCDLA+log_LBXHXCLA,
                         expnms=Xnm,
                         wqs_data, family=gaussian(), q=4)
gcompmod2 <- qgcomp.boot(y~sex+log_LBX074LA+log_LBX099LA+log_LBX105LA+log_LBX118LA+log_LBX138LA+log_LBX153LA+
                           log_LBX156LA+log_LBX157LA+log_LBX167LA+log_LBX170LA+log_LBX180LA+log_LBX187LA+log_LBX189LA+
                           log_LBX194LA+log_LBX196LA+log_LBX199LA+log_LBXD01LA+log_LBXD02LA+log_LBXD03LA+log_LBXD04LA+
                           log_LBXD05LA+log_LBXD07LA+log_LBXF01LA+log_LBXF02LA+log_LBXF03LA+log_LBXF04LA+log_LBXF05LA+
                           log_LBXF06LA+log_LBXF07LA+log_LBXF08LA+log_LBXF09LA+log_LBXPCBLA+log_LBXTCDLA+log_LBXHXCLA,
                         expnms=Xnm,
                         wqs_data, family=gaussian(), q=4, B=200)
wqsmod <- gwqs(y ~ sex, mix_name = Xnm, data = wqs_data, q = 4, 
     validation = 0.6, b = 3, b1_pos = TRUE, b1_constr = F, family='gaussian', seed=125, plots=FALSE)
     

```

From the first plot we see weights from `qgcomp.noboot` function, which include both
positive and negative effect directions. Using `qgcomp.boot` also allows us to assess
linearity of the total exposure effect

```{r, results='markup', message=FALSE, fig.height=7, fig.width=6}
plot(gcompmod)
plot(gcompmod2)
```

Compare with the `gwqs` output plots, which allows effects only in a single direction

```{r, echo=F, results='markup', message=FALSE, fig.height=7, fig.width=6}
wqsmod <- gwqs(y ~ sex, mix_name = Xnm, data = wqs_data, q = 4, 
     validation = 0.6, b = 3, b1_pos = TRUE, b1_constr = F, family='gaussian', seed=125, plots=TRUE)

```
Note the following: the plot for `gcompmod2` (using g-computation with bootstrap variance) gives predictions at the joint intervention levels of exposure. In contrast, the plot `gwqs` gives predictions at levels of the exposure "index" which is a weighted sum of all observed exposures, where the weights are estimated via constrained regression in `qwqs`. Notably, at the lower tail of the `qwqs` plot, the 'index' is populated by observations with exposures corresponding to low weights, meaning they have little contribution to the mean outcome (in the positive direction). In contrast, the lower tail of the `gcompmod2` is based on predicted risk at low joint exposures across all individuals. The two smoothed lines should not, in general, be expected to overlap. The `gcompmod2` plot reflects how population average outcomes are expected to change with joint exposure changes (on the quantile basis), whereas the `qwqs` reflects how average outcomes might change with a change to the exposure index, which is constrained in one direction of effect.

Let's close with one more feature of qgcomp: handling non-linearity. Here is an example where we take the strongest (positive) predictor of the outcome in the primary model (log_LBXD02LA) and fit an indicator variable function

```{r, results='markup', fig.show='hold', fig.height=5, fig.width=5, cache=TRUE}

gcompmod3 <- qgcomp(y~sex+
                           I(log_LBXD02LA==1) + I(log_LBXD02LA==2) + I(log_LBXD02LA==3) +
                           log_LBX074LA+log_LBX099LA+log_LBX105LA+log_LBX118LA+log_LBX138LA+log_LBX153LA+
                           log_LBX156LA+log_LBX157LA+log_LBX167LA+log_LBX170LA+log_LBX180LA+log_LBX187LA+log_LBX189LA+
                           log_LBX194LA+log_LBX196LA+log_LBX199LA+log_LBXD01LA+log_LBXD03LA+log_LBXD04LA+
                           log_LBXD05LA+log_LBXD07LA+log_LBXF01LA+log_LBXF02LA+log_LBXF03LA+log_LBXF04LA+log_LBXF05LA+
                           log_LBXF06LA+log_LBXF07LA+log_LBXF08LA+log_LBXF09LA+log_LBXPCBLA+log_LBXTCDLA+log_LBXHXCLA,
                         expnms=Xnm,
                         wqs_data, family=gaussian(), q=4, B=100)
plot(gcompmod3)
```

Note that allowing for a non-linear effect of log_LBXD02LA induces an apparent non-linear trend in the overall exposure effect. The smoothed regression line is still well within the prediction intervals of the marginal linear model (by default, the overall effect of joint exposure is assumed linear, though this assumption can be relaxed via the 'degree' parameter in qgcomp.boot, as follows:

```{r, results='markup', fig.show='hold', fig.height=5, fig.width=5, cache=TRUE}

gcompmod4 <- qgcomp(y~sex+
                           I(log_LBXD02LA==1) + I(log_LBXD02LA==2) + I(log_LBXD02LA==3) +
                           log_LBX074LA+log_LBX099LA+log_LBX105LA+log_LBX118LA+log_LBX138LA+log_LBX153LA+
                           log_LBX156LA+log_LBX157LA+log_LBX167LA+log_LBX170LA+log_LBX180LA+log_LBX187LA+log_LBX189LA+
                           log_LBX194LA+log_LBX196LA+log_LBX199LA+log_LBXD01LA+log_LBXD03LA+log_LBXD04LA+
                           log_LBXD05LA+log_LBXD07LA+log_LBXF01LA+log_LBXF02LA+log_LBXF03LA+log_LBXF04LA+log_LBXF05LA+
                           log_LBXF06LA+log_LBXF07LA+log_LBXF08LA+log_LBXF09LA+log_LBXPCBLA+log_LBXTCDLA+log_LBXHXCLA,
                         expnms=Xnm,
                         wqs_data, family=gaussian(), q=4, B=100, degree=2)
plot(gcompmod4)
```
Note the prediction interval has been replaced by actual model predictions. Ideally, the smooth fit will look very similar to the model prediction regression line.

## References


## Acknowledgements

The development of this package was supported by NIH Grant RO1ES02953101
