---
title: "How to use qgcomp package"
author: "Alexander Keil"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use qgcomp package}
  %\VignetteEngine{knitr::knitr}
  \usepackage[utf8]{inputenc}
---

## Introduction

## How to use the `qgcomp` package

### Example 1

Here are some observations from  `wqs_data` dataset from the from the package `gQWS`. 
These data will be used as an ongoing example to demonstrate features of the
`qgcomp` package. Parallel examples can be seen in the `gQWS` help files.

```{r, echo=FALSE, results='markup', message=FALSE}
library(gWQS)
library(qgcomp)
library(knitr)
head(wqs_data[, c(37, 36, 35, 1:34)], 10)
```

qgcomp with a continuous outcome:
This script calls a qgcomp model for a continuous outcome using the function `qgcomp.noboot`.
Compare with call to `gqws` from the package `gQWS`.

```{r, results='markup', fig.show='hold', fig.height=5, fig.width=5, cache=TRUE}

# we save the names of the mixture variables in the variable "Xnm"
data("wqs_data", package="gWQS")

Xnm <- c("log_LBX074LA", "log_LBX099LA", "log_LBX105LA", "log_LBX118LA",
"log_LBX138LA", "log_LBX153LA", "log_LBX156LA", "log_LBX157LA", "log_LBX167LA",
"log_LBX170LA", "log_LBX180LA", "log_LBX187LA", "log_LBX189LA", "log_LBX194LA",
"log_LBX196LA", "log_LBX199LA", "log_LBXD01LA", "log_LBXD02LA", "log_LBXD03LA",
"log_LBXD04LA", "log_LBXD05LA", "log_LBXD07LA", "log_LBXF01LA", "log_LBXF02LA",
"log_LBXF03LA", "log_LBXF04LA", "log_LBXF05LA", "log_LBXF06LA", "log_LBXF07LA",
"log_LBXF08LA", "log_LBXF09LA", "log_LBXPCBLA", "log_LBXTCDLA", "log_LBXHXCLA")


# we run the model and save the results in the variable "results"
results <- qgcomp.noboot(y~.,dat=wqs_data[,c(Xnm, 'y')], family=gaussian())


# we compare a qgcomp.noboot fit:
gcompmod <- qgcomp.noboot(disease_state~., expnms=Xnm, data = wqs_data[,c(Xnm, 'disease_state')], family=binomial(), q=4)

# and a qgcomp.boot fit:
gcompmod2 <- qgcomp.boot(disease_state~., expnms=Xnm, data = wqs_data[,c(Xnm, 'disease_state')], family=binomial(), q=4, B=200)


# with a gwqs fit:
suppressWarnings(wqsmod <- gwqs(disease_state ~ 1, mix_name = Xnm, data = wqs_data, q = 4, validation = 0.6, b = 3, b1_pos = F, b1_constr = F, family='binomial', seed=125))
     

# WQS fit (with reduced number of bootstraps to save time)
wqsmod$final_weights
wqsmod$fit
# conditional OR fit, similar to WQS
gcompmod
# population average RR with bootstrap confidence intervals
gcompmod2


```



The `qgcomp.noboot` function gives back other outputs...

### Example 2

In the following code we run an adjusted linear regression (`family = "gaussian"`)...

```{r, results='markup', fig.show='hold', fig.height=5, fig.width=5, cache=TRUE}

gcompmod <- qgcomp.noboot(y~sex+log_LBX074LA+log_LBX099LA+log_LBX105LA+log_LBX118LA+log_LBX138LA+log_LBX153LA+
                           log_LBX156LA+log_LBX157LA+log_LBX167LA+log_LBX170LA+log_LBX180LA+log_LBX187LA+log_LBX189LA+
                           log_LBX194LA+log_LBX196LA+log_LBX199LA+log_LBXD01LA+log_LBXD02LA+log_LBXD03LA+log_LBXD04LA+
                           log_LBXD05LA+log_LBXD07LA+log_LBXF01LA+log_LBXF02LA+log_LBXF03LA+log_LBXF04LA+log_LBXF05LA+
                           log_LBXF06LA+log_LBXF07LA+log_LBXF08LA+log_LBXF09LA+log_LBXPCBLA+log_LBXTCDLA+log_LBXHXCLA,
                         expnms=Xnm,
                         wqs_data, family=gaussian(), q=4)
gcompmod2 <- qgcomp.boot(y~sex+log_LBX074LA+log_LBX099LA+log_LBX105LA+log_LBX118LA+log_LBX138LA+log_LBX153LA+
                           log_LBX156LA+log_LBX157LA+log_LBX167LA+log_LBX170LA+log_LBX180LA+log_LBX187LA+log_LBX189LA+
                           log_LBX194LA+log_LBX196LA+log_LBX199LA+log_LBXD01LA+log_LBXD02LA+log_LBXD03LA+log_LBXD04LA+
                           log_LBXD05LA+log_LBXD07LA+log_LBXF01LA+log_LBXF02LA+log_LBXF03LA+log_LBXF04LA+log_LBXF05LA+
                           log_LBXF06LA+log_LBXF07LA+log_LBXF08LA+log_LBXF09LA+log_LBXPCBLA+log_LBXTCDLA+log_LBXHXCLA,
                         expnms=Xnm,
                         wqs_data, family=gaussian(), q=4, B=200)
wqsmod <- gwqs(y ~ sex, mix_name = Xnm, data = wqs_data, q = 4, 
     validation = 0.6, b = 3, b1_pos = TRUE, b1_constr = F, family='gaussian', seed=125, plots=FALSE)
     

```

From the first plot we see weights from `qgcomp.noboot` function, which include both
positive and negative effect directions. Using `qgcomp.boot` also allows us to assess
linearity of the total exposure effect

```{r, results='markup', message=FALSE, fig.height=7, fig.width=6}
plot(gcompmod)
plot(gcompmod2)
```

Compare with the `gwqs` output plots, which allows effects only in a single direction

```{r, echo=F, results='markup', message=FALSE, fig.height=7, fig.width=6}
wqsmod <- gwqs(y ~ sex, mix_name = Xnm, data = wqs_data, q = 4, 
     validation = 0.6, b = 3, b1_pos = TRUE, b1_constr = F, family='gaussian', seed=125, plots=TRUE)

```


## References


## Acknowledgements

The development of this package was supported by NIH Grant RO1ES02953101
