---
title: "How to use qgcomp package"
author: "Alexander Keil"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use qgcomp package}
  %\VignetteEngine{knitr::knitr}
  \usepackage[utf8]{inputenc}
---

## Introduction

## How to use the `qgcomp` package

### Example 1


```{r, echo=FALSE, results='asis', message=FALSE}
library(gWQS)
library(qgcomp)
library(knitr)
knitr::kable(head(wqs_data[, c(37, 36, 35, 1:34)], 10))
```

qgcomp with a continuous outcome:
This script calls a wqs model for a continuous outcome using the function `gwqs`. 

```{r, results='asis', fig.show='hold', fig.height=5, fig.width=5, cache=TRUE}

# we save the names of the mixture variables in the variable "Xnm"
data("wqs_data", package="gWQS")

Xnm = c("log_LBX074LA", "log_LBX099LA", "log_LBX105LA", "log_LBX118LA",
"log_LBX138LA", "log_LBX153LA", "log_LBX156LA", "log_LBX157LA", "log_LBX167LA",
"log_LBX170LA", "log_LBX180LA", "log_LBX187LA", "log_LBX189LA", "log_LBX194LA",
"log_LBX196LA", "log_LBX199LA", "log_LBXD01LA", "log_LBXD02LA", "log_LBXD03LA",
"log_LBXD04LA", "log_LBXD05LA", "log_LBXD07LA", "log_LBXF01LA", "log_LBXF02LA",
"log_LBXF03LA", "log_LBXF04LA", "log_LBXF05LA", "log_LBXF06LA", "log_LBXF07LA",
"log_LBXF08LA", "log_LBXF09LA", "log_LBXPCBLA", "log_LBXTCDLA", "log_LBXHXCLA")


# we run the model and save the results in the variable "results"
results = qgcomp.noboot(y~.,dat=wqs_data[,c(Xnm, 'y')], family=gaussian())


# we compare...

gcompmod = qgcomp.noboot(disease_state~.,wqs_data[,c(Xnm, 'disease_state')], family=binomial(), q=4)

wqsmod = gwqs(disease_state ~ 1, mix_name = Xnm, data = wqs_data, q = 4, 
     validation = 0.6, b = 3, b1_pos = F, b1_constr = F, family='binomial', seed=125)
     

wqsmod$final_weights
summary(wqsmod$fit)
gcompmod


```

This model ...


```{r, echo=FALSE, results='asis', message=FALSE}
knitr::kable(results$pweights, digits = 3, row.names = F)
knitr::kable(results$nweights, digits = 3, row.names = F)
```

In the second plot...

To see the estimates of the mixture effect, we run the following

```{r, results='asis', message=FALSE, eval=F}
print(results)
```

This last table tells us...  




The `qgcomp.noboot` function gives back other outputs...

### Example 2

In the following code we run a logistic regression (`family = "binomial"`)...

```{r, results='asis', fig.show='hold', fig.height=5, fig.width=5, cache=TRUE}

gcompmod = qgcomp.noboot(disease_state~sex+log_LBX074LA+log_LBX099LA+log_LBX105LA+log_LBX118LA+log_LBX138LA+log_LBX153LA+
                           log_LBX156LA+log_LBX157LA+log_LBX167LA+log_LBX170LA+log_LBX180LA+log_LBX187LA+log_LBX189LA+
                           log_LBX194LA+log_LBX196LA+log_LBX199LA+log_LBXD01LA+log_LBXD02LA+log_LBXD03LA+log_LBXD04LA+
                           log_LBXD05LA+log_LBXD07LA+log_LBXF01LA+log_LBXF02LA+log_LBXF03LA+log_LBXF04LA+log_LBXF05LA+
                           log_LBXF06LA+log_LBXF07LA+log_LBXF08LA+log_LBXF09LA+log_LBXPCBLA+log_LBXTCDLA+log_LBXHXCLA,
                         expcoefs=c(0, rep(1, 34)),
                         wqs_data[,c(Xnm, "sex", 'disease_state')], family=binomial(), q=4)
wqsmod = gwqs(disease_state ~ sex, mix_name = Xnm, data = wqs_data, q = 4, 
     validation = 0.6, b = 3, b1_pos = F, b1_constr = F, family='binomial', seed=125)
     

```

From the first plot we see...

```{r, results='asis', message=FALSE, eval=F}
print(gcompmod)
summary(gcompmod$fit)
```

```{r, echo=F, results='asis', message=FALSE}
wqsmod$final_weights
summary(wqsmod$fit)

```


## References


## Acknowledgements

The development of this package was supported by NIH Grant RO1ES02953101
